Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-12 manuscript_agu.tex
Addressing comments

The campaigns contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM 15k, described in detail below (Sections \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

@@

The Vaisala CL51 and CT25K (photos in Fig.\ \ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910 nm and 905 nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4 km (CL51), 7.7 km (CL31 emulation mode with 5 m vertical resolution), and 7.5 km (CT25K). The vertical resolution is 10 m (5 m configurable) in CL51 and 30 m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6 s for CL51 on AA15‐16, 16 s for CT25K on MARCUS and MICRE, 36 s for CL51 on RV \emph{Polarstern}, and about 2.37 s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910 nm are affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910 nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section \ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910 nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.\ \ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

@@

The Lufft CHM 15k (photo in Fig.\ \ref{fig:map}c) is a ceilometer operating at a near-infrared wavelength of 1064 nm. The maximum range is 15.4 km; the vertical resolution is 5 m in the near range (up to 150 m) and 15 m above; the sampling (temporal) resolution is 2 s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section \ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM 15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

Because the analyzed ICON simulation was free-running (years 2021--2024, inclusive), weather and climate oscillations (such as the El Niño--Southern Oscillation) are not expected to be equivalent to reality at the same time and place. To compare with the observations collected during a different time period (years 2010--2021, inclusive), we compared the model output with observations at the same time of year and geographical location, as determined for each data point, such as a lidar profile or a radiosonde launch. In the ALCF, this was done using the \emph{override\_year} option (\url{https://alcf.peterkuma.net/documentation/cli/cmd_model.html}). For radiosonde profiles, the same mapping of time from was done. That is, when selecting an equivalent profile from the model, the time of the profile was changed so that the time relative to the start of the year was preserved, but the year was changed to one of the four years available in the model data. Thus, for every radiosonde launch, there were four equivalent model profiles. The geographical location was kept the same.

@@

Radiation calculations presented in the results (Section \ref{sec:results}) were completed such that they always represent daily means in order to be consistent with the CERES SYN1deg data. Therefore, every instantaneous profile in the simulated lidar data was assigned a daily mean radiation value corresponding to the day (in the Coordinated Universal Time; UTC). In turn, the average radiation during the entire voyage or station observation period was calculated as the average of the profile values. In the observed lidar data, the daily mean radiation value was taken from the spatially and temporally co-located CERES SYN1deg data for the day (in UTC). The voyage or station average was calculated in the same way.

@@

Fig.\ \ref{fig:cyclone-stability}a, b show a geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section \ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone, and it is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and the model agree relatively well, there are clear differences, particularly in the highest occurrence rate regions -- over the Amundsen Sea, where the Amundsen Sea low is located \cite{turner2013}, and which is more cyclonic in the reanalysis, and around Cape Adare, which is more cyclonic in ICON. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that the model is free-running.

Fig.\ \ref{fig:cyclone-stability}c, d show a geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section \ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (with respect to our SO partitioning; 50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are prevalent elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilising effect \cite{knight2024}. The ERA5 reanalysis is also substantially more stable than ICON across the whole region.

@@

When subsetted by conditions of weak and strong stability (Fig.\ \ref{fig:cloud-occurrence}f, g), as defined in Section \ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly nearly at the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1 km. It is worth mentioning that conditions of strong stability might be associated with the formation of advection fog, such as in situations of warm air advection from the north over colder sea surface thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak by up to 12\%, underestimated cloud occurrence above 1 km by up to 5\%, and underestimated cloud occurrence near the surface by about 11\%. In situations of weak stability, the bias in ICON is very pronounced, with a much stronger peak at about 500 m; ERA5 underestimates cloud occurrence below 1 km (especially near the surface); and MERRA-2 underestimates cloud occurrence even more strongly.

@@

In Fig.\ \ref{fig:cloud-cover} we show the results for the same subsets of data as in Section \ref{sec:cloud-occurrence}. Observations display the highest proportion of high cloud cover values (5--8 oktas), peaking at 7 oktas. This pattern is not represented by ICON or either reanalysis. While ICON is closest to matching the observed distribution, it tends to be 1 okta clearer than the observations, peaking at 6 oktas, and substantially underestimating days with 8 oktas. Overall, the reanalyses show results similar to each other, underestimating cloud cover by about 2 oktas and strongly underestimating days with 7 and 8 oktas. Of the two reanalyses, MERRA-2 has slightly higher cloud cover than ERA5, by about 6\% at 6 octas, which makes it more consistent with observations.

@@

In order to link the cloud biases to the local physical conditions, we analyzed about 2300 radiosonde profiles south of 40°S from the 24 RV \emph{Polarstern} voyages, MARCUS, NBP1704, TAN1702, and TAN1802. Spatially and temporally colocated profiles were taken from ICON and the reanalyses. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with the model, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year of 2021--2024) for each observed profile. The profiles were partitioned into the same subsets as above (Sections \ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). Apart from relative humidity, we focus on comparing virtual potential temperature ($\theta_v$) due to its role in low-level tropospheric stability, being one of the primary factors affecting shallow convection and the associated low-level cloud formation and dissipation. The observed and model profiles of virtual potential temperature are shown in Fig.\ \ref{fig:potential-temperature}.

@@

\section{Limitations of this study}

@@

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations such as the ENSO in the observations and the model.

@@

We have attempted to remove lidar profiles with precipitation, which could not be properly simulated with the lidar simulator (Section \ref{sec:ann}). However, the approach was limited by the relatively low sensitivity of the ANN (65\%) and the fact that we had to choose a fixed threshold for surface precipitation flux in the model and reanalyses, which might not correspond to detection by the ANN applied to observations. Also, we made no attempt to remove profiles with precipitation not reaching the surface. The above reasons may result in an artificial bias in the comparison, even though we expect this to be much smaller than the identified model biases.

>>>>

The campaigns contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM 15k, described in detail below (Sections\ \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

@@

The Vaisala CL51 and CT25K (photos in Fig.\ \ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910 nm and 905 nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4 km (CL51), 7.7 km (CL31 emulation mode with 5 m vertical resolution), and 7.5 km (CT25K). The vertical resolution is 10 m (5 m configurable) in CL51 and 30 m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6 s for CL51 on AA15‐16, 16 s for CT25K on MARCUS and MICRE, 36 s for CL51 on RV \emph{Polarstern}, and about 2.37 s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910 nm are affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910 nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section\ \ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910 nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.\ \ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

@@

The Lufft CHM 15k (photo in Fig.\ \ref{fig:map}c) is a ceilometer operating at a near-infrared wavelength of 1064 nm. The maximum range is 15.4 km; the vertical resolution is 5 m in the near range (up to 150 m) and 15 m above; the sampling (temporal) resolution is 2 s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section\ \ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM 15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

Because the analyzed ICON simulation was free-running (years 2021--2024, inclusive), weather and climate oscillations (such as the El Niño--Southern Oscillation) are not expected to be equivalent to reality at the same time and place. To compare with the observations collected during a different time period (years 2010--2021, inclusive), we compared the model output with observations at the same time of year and geographical location, as determined for each data point, such as a lidar profile or a radiosonde launch. In the ALCF, this was done using the \emph{override\_year} option (\url{https://alcf.peterkuma.net/documentation/cli/cmd_model.html}). For radiosonde profiles, the same mapping of time from was done. That is, when selecting an equivalent profile from the model, the time of the profile was changed so that the time relative to the start of the year was preserved, but the year was changed to one of the four years available in the model data. Thus, for every radiosonde launch, there were four equivalent model profiles. The geographical location was kept the same. We discuss briefly the implications of comparing the observations with a free-running model in Section\ \ref{sec:limitations}.

@@

Radiation calculations presented in the results (Section\ \ref{sec:results}) were completed such that they always represent daily means in order to be consistent with the CERES SYN1deg data. Therefore, every instantaneous profile in the simulated lidar data was assigned a daily mean radiation value corresponding to the day (in the Coordinated Universal Time; UTC). In turn, the average radiation during the entire voyage or station observation period was calculated as the average of the profile values. In the observed lidar data, the daily mean radiation value was taken from the spatially and temporally co-located CERES SYN1deg data for the day (in UTC). The voyage or station average was calculated in the same way.

@@

Fig.\ \ref{fig:cyclone-stability}a, b show a geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section\ \ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone, and it is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and the model agree within about 8\% in most areas, ICON is prevailingly more cyclonic by about 4\%. There are clear differences particularly in the highest occurrence rate regions such as around Cape Adare, which is by up to 20\% more cyclonic in ICON, and Weddel and Bellingshausen Sea, where ICON is less cyclonic by up to 10\%. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that the model is free-running.

Fig.\ \ref{fig:cyclone-stability}c, d show a geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section\ \ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (with respect to our SO partitioning; 50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are prevalent elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilising effect \cite{knight2024}. The ERA5 reanalysis is also substantially more stable than ICON across the whole region.

@@

When subsetted by conditions of weak and strong stability (Fig.\ \ref{fig:cloud-occurrence}f, g), as defined in Section\ \ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly nearly at the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1 km. It is worth mentioning that conditions of strong stability might be associated with the formation of advection fog, such as in situations of warm air advection from the north over colder sea surface thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak by up to 12\%, underestimated cloud occurrence above 1 km by up to 5\%, and underestimated cloud occurrence near the surface by about 11\%. In situations of weak stability, the bias in ICON is very pronounced, with a much stronger peak at about 500 m; ERA5 underestimates cloud occurrence below 1 km (especially near the surface); and MERRA-2 underestimates cloud occurrence even more strongly.

@@

In Fig.\ \ref{fig:cloud-cover} we show the results for the same subsets of data as in Section\ \ref{sec:cloud-occurrence}. Observations display the highest proportion of high cloud cover values (5--8 oktas), peaking at 7 oktas. This pattern is not represented by ICON or either reanalysis. While ICON is closest to matching the observed distribution, it tends to be 1 okta clearer than the observations, peaking at 6 oktas, and substantially underestimating days with 8 oktas. Overall, the reanalyses show results similar to each other, underestimating cloud cover by about 2 oktas and strongly underestimating days with 7 and 8 oktas. Of the two reanalyses, MERRA-2 has slightly higher cloud cover than ERA5, by about 6\% at 6 octas, which makes it more consistent with observations.

@@

In order to link the cloud biases to the local physical conditions, we analyzed about 2300 radiosonde profiles south of 40°S from the 24 RV \emph{Polarstern} voyages, MARCUS, NBP1704, TAN1702, and TAN1802. Spatially and temporally colocated profiles were taken from ICON and the reanalyses. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with the model, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year of 2021--2024) for each observed profile. The profiles were partitioned into the same subsets as above (Sections\ \ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). Apart from relative humidity, we focus on comparing virtual potential temperature ($\theta_v$) due to its role in low-level tropospheric stability, being one of the primary factors affecting shallow convection and the associated low-level cloud formation and dissipation. The observed and model profiles of virtual potential temperature are shown in Fig.\ \ref{fig:potential-temperature}.

@@

\section{Limitations of this study}
\label{sec:limitations}

@@

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations such as the ENSO in the observations and the model. The interannual variability in cloud occurrence in ICON can be seen in Fig.\ \ref{fig:cloud-occurrence-panel}, where each year in ICON is represented by a separate line. The interannual variability tends to be substantially smaller than the biases, and thus is unlikely to have a stong impact on the main findings.

@@

We have attempted to remove lidar profiles with precipitation, which could not be properly simulated with the lidar simulator (Section\ \ref{sec:ann}). However, the approach was limited by the relatively low sensitivity of the ANN (65\%) and the fact that we had to choose a fixed threshold for surface precipitation flux in the model and reanalyses, which might not correspond to detection by the ANN applied to observations. Also, we made no attempt to remove profiles with precipitation not reaching the surface. The above reasons may result in an artificial bias in the comparison, even though we expect this to be much smaller than the identified model biases.
